Gravitational waves from colliding black holes

### 2017-05-04 Report/Draft/draft.tex
Report
Report latest

\begin{document}
\maketitle
\begin{abstract}
A complete understanding of the LIGO detectors' ability to accurately measure spin parameters is of key importance to understanding the astrophysics of binary black hole systems. Of particular significance is an ability to determine spin orientations as a means of distinguishing between binary formation models. In this work we overview the core physics of gravitational waves, relativistic precession in binary systems, the LIGO detectors, and the techniques involved in parameter estimation. We then examine the specific effects of precession on a gravitational waveform. The technique of 'matching' between waveforms is used as a computationally cheap way to explore degeneracies in the parameter space. A series of signals are then injected into simulated detector noise, and a full parameter estimation is performed on the data segments. Using the match results as a guide, we target these injections to areas in the parameter space where we expect to be particularly sensitive and insensitive to spin parameters. Inference for a full range of inclinations and phases is performed, and we identify areas where spin orientations are consistently incorrectly recovered. We then briefly examine the impact the upcoming VIRGO detector will have on spin estimation.
\end{abstract}
\tableofcontents
\newpage

@@

In this section we present the results from a range of inference runs, attempt to assess the effectiveness of spin estimation and compare the inference results with the match results to evaluate how effective the match process is as a predictor of inference accuracy. For all injections we set $S_{1\perp}=0.9$ and $|\vec{S_2}|=0$ with $\chi_p=0.9$, and cover a uniform range of inclinations. We set the phase to be $\phi=0.$, and the polarisation angle $\psi=0.8$, so the signals are of mixed polarisation, but slightly cross dominated. The match results indicate that cross polarised signals generally have a higher spin resolution, but we want to avoid the signal drop at $\iota=\pi/2$, which is why a mixed but cross dominated polarisation angle was selected. The source location and coalesence time were also kept fixed for all injections. We also only examine high mass ratio binaries, as the match results indicate that precession is not significant in equal mass ratio binaries. Prior distributions are all uniform, with mass ranging $[5,80]$ and distance in the range $[100,1000]$, and sky location and spin orientations all have isotropic priors. Out of a total of ****** software injections that were performed, there were only **** instances where the injected $\chi_p$ value fell within the 90\% confidence interval. In Figs 14 and 15 we compare posteriors for two inferences, where $\chi_p$ was excluded and where it was recovered moderately well, and we refer to these as A and B respectively. Even in the cases where the injected value was within the 90\% confidence intervals, it was always closer to the upper bound than to the mean, and a mean $\chi_p$ greater than 0.65 was never found.

@@

From Fig. 18 it can be seen that the amplitude of the signal noticeably drops towards the end of the inspiral and just before the merger, implying that this would be an ideal signal to recover a high value of $\chi_p$ from. The posteriors shown in Fig. 17 are mostly accurate, with the majority of injected parameters lying inside the 90\% confidence intervals, however once again $\chi_p$ is excluded, and significantly underestimated. The match values shown in Fig. 18 are strong, with both detectors giving a match of 0.96 between the injected and MAP waveforms. This is further evidence of an intrinsic degeneracy that prevents high $\chi_p$ MAP values from being found. For all posteriors shown, both $\phi$ and $\psi$ are broadly uniform, and in the case where they are not uniform, they are largely inaccurate. However in section 2 it was observed that $\psi$ and especially $\phi$ have a significant effect on the structure of the parameter space, and the match between precessing and non-precessing waveforms.

@@

Next we examine the effect of inclination on $\chi_p$ recovery. Software injections were performed for a range of inclinations at two distances, 200Mpc and 800Mpc, which simulate a strong and weak signal respectively. Posteriors for $\chi_p$ and $\chi_{eff}$ are shown in Fig. 19 for 6 inclinations in the range $[0,\pi]$, with the injected values shown by the red bar. Interestingly there is no apparent correlation between the trends for $\chi_p$ at 200Mpc and at 800Mpc. If anything the trends appear to be opposite, with the most greater accuracy at face-on for the signals at 200Mpc, and greatest accuracy close to edge-on for the 800Mpc signals. As in the previous results, $\chi_{eff}$ is recovered consistently well with the exception of two results around $\iota\approx2$ in the 200Mpc signals, again implying that spin orientation is accurate, but magnitude is underestimated. Therefore there appears to be no clear correlation between $\chi_p$ and $\chi_{eff}$ accuracy.
\begin{figure}[h]
	\includegraphics[width=1\textwidth]{fig19.png}
	\centering
	\caption{Comparisons of derived spin parameters as a function inclination for strong and weak signals. All other parameters are kept the same as listed in Fig. 14.}
	\centering
\end{figure}
\begin{figure}[h]
	\includegraphics[scale=0.6]{fig20.png}
	\centering
	\caption{Match between MAP and injected waveforms along with spin inaccuracy for the 200Mpc and 800Mpc runs as a function of inclination.}
	\centering
\end{figure}

In Fig. 20 we attempt to identify how well the posteriors in Fig. 19 correlate with the predicted trends from the match results in Figs. 10 and 11. If the match process were a reliable method of predicting inference results, the posteriors for $\chi_p$ around inclinations of 1 to 2 should be the most accurate, as precessing waveforms are most distinct from precessing ones here. However the 200Mpc results in fact show the opposite trend, and accuracy drops to a minimum around this area. 
\subsection{Impact of Virgo}
A look at how Virgo will influence PE, especially chi p estimation.
\begin{figure}[h]
	\includegraphics[width=1\textwidth]{fig21.png}
	\centering
	\caption{Plots showing $\chi_p$ and $\chi_{eff}$ for a range of inclinations at 200Mpc with the VIRGO detector included in the inference process.}
	\centering
\end{figure}
\section{Conclusions}
Summary of results on PE and estimation of chi p, and a prospect on Virgo's impact.
\section{Reflective Statement}
I now need a therapist

Still to do:

6. Conclusions
7. Proper references of pics

>>>>

\begin{document}
\maketitle
\begin{abstract}
A complete understanding of the LIGO detectors' ability to accurately measure spin parameters is of key importance to understanding the astrophysics of binary black hole systems. Of particular significance is an ability to determine spin orientations as a means of distinguishing between binary formation models. In this work we overview the core physics of gravitational waves, relativistic precession in binary systems, the LIGO detectors, and the techniques involved in parameter estimation. We then examine the specific effects of precession on a gravitational waveform. The technique of 'matching' between waveforms is used as a computationally cheap way to explore degeneracies in the parameter space. A series of signals are then injected into simulated detector noise, and a full parameter estimation is performed on the data segments. Using the match results as a guide, we target these injections to areas in the parameter space where we expect to be particularly sensitive and insensitive to spin parameters. Inference for a full range of inclinations is performed, and we identify areas where spin orientations are consistently incorrectly recovered. We then briefly examine the impact the upcoming VIRGO detector will have on spin estimation.
\end{abstract}
\tableofcontents
\newpage

@@

In this section we present the results from a range of inference runs, attempt to assess the effectiveness of spin estimation and compare the inference results with the match results to evaluate how effective the match process is as a predictor of inference accuracy. For all injections we set $S_{1\perp}=0.9$ and $|\vec{S_2}|=0$ with $\chi_p=0.9$, and cover a uniform range of inclinations. We set the phase to be $\phi=0.$, and the polarisation angle $\psi=0.8$, so the signals are of mixed polarisation, but slightly cross dominated. The match results indicate that cross polarised signals generally have a higher spin resolution, but we want to avoid the signal drop at $\iota=\pi/2$, which is why a mixed but cross dominated polarisation angle was selected. The source location and coalesence time were also kept fixed for all injections. We also only examine high mass ratio binaries, as the match results indicate that precession is not significant in equal mass ratio binaries. Prior distributions are all uniform, with mass ranging $[5,80]$ and distance in the range $[100,1000]$, and sky location and spin orientations all have isotropic priors. Out of a total of 20 software injections that were performed, there were only 5 instances where the injected $\chi_p$ value fell within the 90\% confidence interval. In Figs 14 and 15 we compare posteriors for two inferences, where $\chi_p$ was excluded and where it was recovered moderately well, and we refer to these as A and B respectively. Even in the cases where the injected value was within the 90\% confidence intervals, it was always closer to the upper bound than to the mean, and a mean $\chi_p$ greater than 0.65 was never found.

@@

From Fig. 18 it can be seen that the waveform shows strong modulation, with the amplitude of the signal dropping significantly towards the end of the inspiral and just before the merger, implying that this would be an ideal signal to recover a high value of $\chi_p$ from. The posteriors shown in Fig. 17 are mostly accurate, with the majority of injected parameters lying inside the 90\% confidence intervals, however once again $\chi_p$ is excluded, and significantly underestimated. The match values shown in Fig. 18 are strong, with both detectors giving a match of 0.96 between the injected and MAP waveforms. This is further evidence of an intrinsic degeneracy that prevents high $\chi_p$ MAP values from being found. For all posteriors shown, both $\phi$ and $\psi$ are broadly uniform, and in the case where they are not uniform, they are largely inaccurate. However in section 2 it was observed that $\psi$ and especially $\phi$ have a significant effect on the structure of the parameter space, and the match between precessing and non-precessing waveforms.

@@

Next we examine the effect of inclination on $\chi_p$ recovery. Software injections were performed for a range of inclinations at two distances, 200Mpc and 800Mpc, which simulate a strong and weak signal respectively. Posteriors for $\chi_p$ and $\chi_{eff}$ are shown in Fig. 19 for 6 inclinations in the range $[0,\pi]$, with the injected values shown by the red bar. Interestingly there is no apparent correlation between the trends for $\chi_p$ at 200Mpc and at 800Mpc. If anything the trends appear to be opposite, with the most greater accuracy at face-on for the signals at 200Mpc, and greatest accuracy close to edge-on for the 800Mpc signals. As in the previous results, $\chi_{eff}$ is recovered consistently well with the exception of two results around $\iota\approx2$ in the 200Mpc signals, again implying that spin orientation is accurate, but magnitude is underestimated. Therefore there appears to be no clear correlation between $\chi_p$ and $\chi_{eff}$ accuracy, although overall the 200Mpc runs are slightly less inaccurate than the 800Mpc signals, likely due to the stronger signal strength.
\begin{figure}[h]
	\includegraphics[width=1\textwidth]{fig19.png}
	\centering
	\caption{Comparisons of derived spin parameters as a function inclination for strong and weak signals. All other parameters are kept the same as listed in Fig. 14.}
	\centering
\end{figure}
\begin{figure}[h]
	\includegraphics[scale=0.6]{fig20.png}
	\centering
	\caption{Match between MAP and injected waveforms along with spin inaccuracy for the 200Mpc and 800Mpc runs as a function of inclination.}
	\centering
\end{figure}

In Fig. 20 we attempt to identify how well the posteriors in Fig. 19 confirm the predicted trends from the match results in Figs. 10 and 11. This is done by plotting the match between the MAP and injected waveforms for each detector, along with the $\chi_p$ inaccuracy. If the match process were a reliable method of predicting inference results, the posteriors for $\chi_p$ around inclinations of 1 to 2 should be the most accurate, as precessing waveforms are most distinct from precessing ones here. So we would expect the spin inaccuracies to drop to a minimum around the face-on inclination. However the 200Mpc results in fact show the opposite trend, and inaccuracy of the 200Mpc signals actually peaks around this area. The 800Mpc shows a different trend, with the $\chi_p$ inacurracy being roughly constant, except from one result at $\iota=1.9$. These results imply that the matching process does not reliably predict inference results, at least for $\chi_p$ estimation. Fig 20. shows the match between the MAP and injected waveforms is strongest at edge-on, implying the inference process is working most effectively for these inclinations. The other key result though is that the spin inaccuracy and matches shown in Fig 20 are not clearly correlated with each other. This again indicates that even when the inference process recovers a MAP waveform with a strong match with the injected waveform, it is not necessarily a waveform that faithfully represents the parameters of the original signal due to the abundance of degeneracies. 
\subsection{Impact of Virgo}
Finally we briefly look toward the influence of the upcoming Virgo detector on the recovery of spin parameters, particularly $\chi_p$. The main contribution Virgo is expected to add to the network is dramatically improved source location due to an added trianglulation effect, however it is also interesting to consider whether it will have any dramatic effect on recovery of other parameters. Results for a range of signals injected at 800Mpc with all parameters kept the same, except with Virgo now included in the inference process, are shown in Fig. 21. The $\chi_p$ results are once again consistently inaccurate, with the injected value being exluced for every run, in fact making this series of injections the least accurate overall. However in this case, the posteriors for $\chi_p$ show a clearer trend as a function of inclination, with the posteriors around face-on being pulled up towards the injected value. So given these results, it appears that adding Virgo to the network makes the inference results slightly more consistent, but no more accurate. The results for $\chi_{eff}$ are almost identical to the runs without Virgo, which indicates that Virgo is unlikely to have any significant impact on spin estimation.
\begin{figure}[h]
	\includegraphics[width=1\textwidth]{fig21.png}
	\caption{Plots showing $\chi_p$ and $\chi_{eff}$ for a range of inclinations at 800Mpc with the VIRGO detector included in the inference process.}
This is investigated further in Fig. 21, where we show the detector response to a heavily precessing waveform (with the same parameters as used previously, except with $\psi=2.8$, $\iota=1.89$ and $\phi=1.5$) for all three detectors. The signal, especially the precession modulation is very similar in all three detectors, so adding Virgo to the network does not provide significant additional information to the process that will impact spin paramter recovery. As a binary system precesses during merger, the signal that is modulated in one polarisation essentially leaks into another, at a different inclination. This would mean that at certain combinations of source location and inclination, two detectors oriented at orthogonal polarisations should give ideal $\chi_p$ resolution as the signal modulated in one detector would appear in the other. However the data in Fig. 22 shows that this is not the case for the relative orientations of H1, L1 and Virgo. This is by no means a conclusive study however, as we only consider a fixed phase, polarisation, source location, mass ratio and total mass, and it is easily possible that other configurations of these parameters could lead to different trends.
\begin{figure}[h]
	\includegraphics[width=1\textwidth]{fig22.png}
	\caption{A heavily precessing waveform in three detector frames, H1, L1 and Virgo.}
\section{Conclusions}
Overall the results indicate that the current inference methods consistently underestimate values of $\chi_p$, with all but 5 out of the 20 inference runs excluding the injected value of $\chi_p$. The match results and waveform plots demonstrate that the phase of the GW signal is an important parameter in precessing systems unlike in non-precessing systems, leading to an extra degree of freedom in precessing systems. The matching process was used to highlight the way in which mass ratios, polarisations, inclinations and phases affect precessing waveforms, particularly our ability to distinguish bewteen precessing and non-precessing signals. Degenerate areas of the parameter space were also identified using this technique, and it was identified that near equal mass ratios, precessing and non-precessing waveforms are effectively entirely degenerate. It was also shown that other parameters such as source location, polarisation and inclination affect the way that the modulation effect manifests itself in a specific signal. This leads to a key observation that the strength of signal modulation in a GW waveform is not necessarily proportional to the amount of precession in the binary itself.

The inference pipeline used to perform parameter estimation was then described, and results were compared for a case of $\chi_p$ being recovered, and $\chi_p$ being excluded. Two explanations were proposed for the failure to accurately infer $\chi_p$, one being that the inference method is not working properly, and the other that there are so many intrinsic degeneracies in precessing waveforms that the inference process cannot distinguish between them. A way of distinguishing between these was also proposed, by performing a match between the recovered MAP waveform and the injected signal. This match was consistently high despite the inaccuracies in the posteriors, and so the source of the inaccuracy is likely to be intrinsic degenerecy. If these results are correct, this has significant implications for the future of spin estimation, as even improvements in detector accuracy and inference methodologies are unlikely to be able to make significant advancements in spin recovery if the signals are inherently degenerate. This would also explain the apparent lack of consistency between match results and spin inaccuracy results, as the ability to recover $\chi_p$ appears to be independent of the match between precessing and non-precessing waveforms.
\section{Reflective Statement}
I now need a therapist

7. Proper references of pics
